### 2024-09-06 manuscripts/07-economic-letters-v5/Hadah-Hispanic-Last-Names.tex
paper less than 12 pages and 2000 words

A large literature documents substantial earnings gaps across race and ethnicity \citep{bayer2018divergent, charles2008prejudice, card1992school, fryer2004causes, rubinstein2014pride, bertrand2004emily, juhn1991accounting}. Hispanics constitute a large and growing portion of the population in the United States. As the number of Hispanics increases, determining whether ethnic discrimination influences their labor market outcomes becomes increasingly crucial \citep{chettyUnitedStatesStill2014, chettyEffectsExposureBetter2016,chettyFadingAmericanDream2017,abramitzkyImmigrantsAssimilateMore2020a, abramitzkyNationImmigrantsAssimilation2014,abramitzkyCulturalAssimilationAge2016,chettyWhereLandOpportunity2014}. Thus, it is important to understand how a person's ethnicity affects their labor market outcomes. Assimilation and mobility are crucial because they reflect how well Hispanics can integrate into society and move up the socioeconomic ladder.

In this paper, I answer the following question: does having a Hispanic last name affect education and labor market outcomes? Moreover, I aim to show that comparing Hispanic Whites to non-Hispanic Whites might create an artificially higher earnings gap since the two groups differ on many observable characteristics. \footnote{Observable characteristics refer to factors that can be measured and quantified, such as education level, work experience, and immigration status.} Others have attempted to compare how native-born White Hispanics fare compared to non-Hispanic Whites and foreign-born Hispanics. In \citet{antman2020ethnic,antmanEthnicAttritionObserved2016,antmanEthnicAttritionObserved2016a,antmanEthnicAttritionAssimilation2020}, the authors compare the health and educational outcomes of Hispanic Whites to non-Hispanic Whites and native-born Hispanics to foreign-born Hispanics. They find gaps in education and health between Hispanics and Whites. They also find that native-born Hispanics are more likely than their foreign-born counterparts to report poor health. \citet{davilaChangesRelativeEarnings2008} documents many gaps in labor market outcomes between Hispanics and Whites. They attribute a big part of this gap to differences in education, experience, immigration status, and regional differences.

The US population is growing in diversity. The proportion of non-Whites has increased by more than 10 percentage points from 13 percent in 1995 to 23 percent in 2019. The number of Hispanics has grown by 9 percentage points from 9 percent in 1995 to 16 percent in 2019.\footnote{The portion of non-Whites and Hispanics is calculated using the Current Population Survey (CPS).} Native-born White Hispanic men earn 21\% less than White men, although a substantial portion of the earnings gap is due to educational differences between Hispanics and Whites \citep{duncan2006hispanics, duncan2018identifying, duncan2018socioeconomic}. Some of the earnings differences may also be due to discrimination which will have negative consequences. For example, discrimination against Hispanics can lead to reduced job opportunities, lower wages, and hinder assimilation. In this paper, I examine the role of having a Hispanic last name on labor market and educational outcomes.

Identifying discrimination is difficult due to factors affecting labor market outcomes that are unobservable to economists, such as unobserved skills, and the challenge of separating these factors from prejudice and stereotypes. One strategy used by researchers is audit or resume studies. \citet{bertrand2004emily} conducted an audit study where identical resumes were sent to employers with White and Black-sounding names. This approach, however, has its drawbacks. Audit studies only observe callbacks, not wages.

This study utilizes a method developed by \citet{rubinstein2014pride}. I compare children from inter-ethnic marriages. More precisely, I compare children of Hispanic fathers and White mothers (henceforth HW) to children of White fathers and Hispanic mothers (henceforth  WH). This approach stems from the fact that there is a strong selection among many characteristics, and thus, marriages are not random. Couples match on several observable characteristics like income, schooling, socio-economic background, etc. \citep{averettBetterWorseRelationship2008, averettEconomicRealityBeauty1996, beckerTheoryMarriagePart1973, beckerTheoryMarriagePart1974, beckerTreatiseFamily1993, browningCollectiveUnitaryModels2006, chiapporiFatterAttractionAnthropometric2012}. Children of HW and WH marriages have more similar observable characteristics than children of endogamous/homogamous marriages--- i.e., White fathers-White mothers and Hispanic fathers-Hispanic mothers. Moreover, children from a Hispanic father and White mother household will have a Hispanic last name from their fathers, enabling the investigation of how ethnic signals, such as having a Hispanic last name, affect annual log earnings.

The main identifying assumption of my empirical strategy depends on the assumption that people born to HW parents are similar to their WH peer in all observable---and unobservable---aspects and characteristics that are important in the labor market. Consequently, the only difference between the two groups is the variation in which group is more likely to have a Hispanic sounding last name. Children from mixed ethnic backgrounds may appear physically similar to those from single ethnicity backgrounds, but the influence of family dynamics and upbringing, crucial in developing skills and personal characteristics, varies with the pattern of mixed ethnic marriages. Particularly in a society where Hispanic ancestry is perceived negatively, it raises the question of what kind of White women would choose Hispanic men. Furthermore, even if such unions were formed randomly, children from White-Hispanic homes might benefit from more favorable family conditions than those from Hispanic-White homes, considering Whites generally have stronger socio-economic backgrounds than Hispanics. These factors introduce doubts regarding whether children from Hispanic-White families receive comparable familial support and influences, either genetically or environmentally, as those from White-Hispanic families.

I find that a person with a Hispanic last name receives less education, is more likely to be unemployed, and earns less. The results suggest that individuals with Hispanic last names receive 0.2 years less education and are more likely to be unemployed. The gap in education is equivalent to a 1.9\% reduction, and the unemployment gap is equal to a 14\% increase. The earnings gap between individuals with Hispanic last names and those with White last names is 5 percentage points. The earnings gap between individuals with Hispanic last names and those with White last names is 5 percentage points. However, when controlling for education, the earnings gap reduces to 1 percentage point. The results suggest that the earnings gap is largely due to differences in education.

@@

I use two datasets: the IPUMS Current Population Survey (CPS) Annual Social and Economic (ASEC) \citep{cps2019} and the 1960 to 2000 US censuses \citep{acs2019}.

The CPS data (1994-2019) is used to study the effect of Hispanic last names on labor market outcomes. The census data is employed to construct synthetic parents using the method developed by \citet{rubinstein2014pride}.

\subsection{Children of the four parental types}

The CPS sample is restricted to White, U.S.-born citizens aged 25-40, born between 1960 and 2000. Parents are classified as Hispanic if born in a Spanish-speaking country or Puerto Rico, and White if they are US-born. Four parental types are identified: (1) White father and White mother (WW), (2) White father and Hispanic mother (WH), (3) Hispanic father and White mother (HW), and (4) Hispanic father and Hispanic mother (HH).

Table \ref{tab:combined-mat} shows the distribution of these types. WW children comprise 96\% of the sample, HH 3\%, and inter-ethnic children (WH and HW) 1.35\%. Summary statistics for each group are presented in Table \ref{tab:combined-sum}.

\subsection{Synthetic parents}

Using the 1960-2000 censuses, I constructed a dataset of synthetic parents, including married White men and women born between 1920-1975. Hispanic individuals are defined as White and born in a Spanish-speaking country, while Whites are native-born.

Table \ref{tab:combined-mat} shows the distribution of couple types: WW (96\%), HH (2\%), WH (<1\%), and HW (1\%). Summary statistics for the parents are presented in Table \ref{tab:combined-sum}.

@@

In this section, I present two empirical strategies. The first empirical strategy estimates the effect of having a Hispanic last name on educational outcomes and earnings.

@@

\subsection{Effect of Hispanic Last Name on Educational Outcomes}

Analysis of White, U.S.-born Hispanics ages 25-40 (Table \ref{tab:lastname-ed-reg}) reveals that HW children receive 0.2 years less education than WH children---that represents a 1.5\% reduction. While there is no significant difference in high school dropout rates, HW children are 2 percentage points less likely to earn an associate degree and 3 percentage points less likely to earn a bachelor's degree, which is equal to a 15\% reduction. These gaps are larger for HW women in higher education. These disparities suggest potential barriers or discrimination in access to higher education for HW children, particularly women.

\subsection{Effect on Labor Market Outcomes}

Results from Table \ref{tab:lastnamereg} indicate a 1 percentage point employment gap between HW and WH workers, and a 5 percentage point crude earnings gap. However, when controlling for education, the earnings gap reduces to a statistically insignificant 1 percentage point. I also find that HW children are 1 percentage point more likely to be unemployed than WH children. The gap in unemployment is equivalent to a 14\% increase and is statistically significant even after controlling for education. These results suggest that having a Hispanic last name negatively impacts labor market outcomes, particularly in terms of employment.

@@

This study examines discrimination against Hispanics in the U.S. labor market, focusing on the impact of Hispanic last names and identification on education and earnings. This research finds that inter-ethnic individuals with Spanish-sounding last names receive 0.2 years less education and are more likely to be unemployed. HW children earn 5 percentage points less than WH children, but this gap becomes statistically insignificant when controlling for education.

These results align with existing literature on disparities in educational access \citep{bergman2018education,gaddis2024racial}. While the earnings gap diminishes when controlling for education, this doesn't necessarily indicate an absence of discrimination, as education itself can be influenced by bias. Moreover, educational differences cannot explain the gaps in unemployment, suggesting that discrimination may play a role in labor market outcomes of individuals with Hispanic last names. Further research is needed to comprehensively understand the earnings gaps between Hispanics and Whites, particularly considering that mothers of HW children have higher education and earnings, which theoretically should lead to better outcomes for their children.

>>>>

A substantial earnings gap exists between native-born White Hispanic men and White men, with Hispanic men earning 21\% less. While a significant portion of this gap is attributed to educational differences, discrimination may also play a role, potentially leading to reduced job opportunities, lower wages, and hindered assimilation \citep{duncan2006hispanics, duncan2018identifying, duncan2018socioeconomic}. This paper examines the impact of Hispanic last names on labor market and educational outcomes, contributing to the literature by isolating the effect of ethnicity-signaling surnames from other potentially confounding factors. By focusing on children of inter-ethnic marriages, this study provides a novel approach to understanding the mechanisms behind Hispanic-White earnings disparities and the potential role of name-based discrimination in the labor market.

This study investigates whether having a Hispanic last name affects educational and labor market outcomes. Previous research has identified gaps in education, health, and labor market outcomes between Hispanics and Whites \citep{antman2020ethnic, davilaChangesRelativeEarnings2008}. However, comparing Hispanic Whites to non-Hispanic Whites might artificially inflate the earnings gap. This is due to observable differences in education level, work experience, and immigration status.

Identifying discrimination is challenging due to unobservable factors affecting labor market outcomes. While audit studies provide insights, they primarily measure callback rates rather than full labor market outcomes \citep{bertrand2004emily}. This study compares children from inter-ethnic marriages: those of Hispanic fathers and White mothers (HW) to those of White fathers and Hispanic mothers (WH) \citep{rubinstein2014pride}. This approach leverages marriage selection based on observable characteristics \citep{averettBetterWorseRelationship2008}, allowing for the investigation of how ethnic signals, such as Hispanic last names, affect earnings.

I find that individuals with Hispanic-sounding last names receive 0.2 years less education and are more likely to be unemployed. The gap in education is equivalent to a 1.9\% reduction, and the unemployment gap is equal to a 14\% increase. The earnings gap between individuals with Hispanic-sounding last names and those with White last names is 5 percentage points. However, when controlling for education, the earnings gap reduces to 1 percentage point, suggesting that the earnings gap is largely due to differences in education.

@@

I use two datasets: the IPUMS Current Population Survey (CPS) Annual Social and Economic (ASEC) \citep{cps2019} and the 1960 to 2000 US censuses \citep{acs2019}. The CPS data (1994-2019) is used to study the effect of Hispanic-sounding last name on labor market outcomes. The census data is employed to construct synthetic parents using the method developed by \citet{rubinstein2014pride}.

The CPS sample is restricted to White, U.S.-born citizens aged 25-40, born between 1960 and 2000. Parents are classified as Hispanic if born in a Spanish-speaking country or Puerto Rico, and White if they are US-born. Four parental types are identified: (1) White father and White mother (WW), (2) White father and Hispanic mother (WH), (3) Hispanic father and White mother (HW), and (4) Hispanic father and Hispanic mother (HH). Table \ref{tab:combined-mat} shows the distribution of these types. WW children comprise 96\% of the sample, HH 3\%, and inter-ethnic children (WH and HW) 1.35\%. Summary statistics for each group are presented in Table \ref{tab:combined-sum}. Using the 1960-2000 censuses, I constructed a dataset of synthetic parents, including married White men and women born between 1920-1975. Hispanic individuals are defined as White and born in a Spanish-speaking country, while Whites are native-born. Table \ref{tab:combined-mat} shows the distribution of couple types: WW (96\%), HH (2\%), WH (<1\%), and HW (1\%). Summary statistics for the parents are presented in Table \ref{tab:combined-sum}.

@@

In this section, I present my empirical strategy. The empirical strategy estimates the effect of having a Hispanic last name on educational outcomes and earnings.

@@

My analysis of White, U.S.-born Hispanics ages 25-40 (Table \ref{tab:lastname-ed-reg}) reveals that HW children receive 0.2 years less education than WH children---that represents a 1.5\% reduction. While there is no significant difference in high school dropout rates, HW children are 2 percentage points less likely to earn an associate degree and 3 percentage points less likely to earn a bachelor's degree, which is equal to a 15\% reduction. These gaps are larger for HW women in higher education. These disparities suggest potential barriers or discrimination in access to higher education for HW children, particularly women.

I show in Table \ref{tab:lastnamereg} a 1 percentage point employment gap between HW and WH workers, and a 5 percentage point crude earnings gap. However, when controlling for education, the earnings gap reduces to a statistically insignificant 1 percentage point. I also find that HW children are 1 percentage point more likely to be unemployed than WH children. The gap in unemployment is equivalent to a 14\% increase and is statistically significant even after controlling for education. These results suggest that having a Hispanic last name negatively impacts labor market outcomes, particularly in terms of employment.

@@

This study examines discrimination against Hispanics in the U.S. labor market, focusing on the impact of Hispanic-sounding last name and identification on education and earnings. This research finds that inter-ethnic individuals with Spanish-sounding last names receive 0.2 years less education and are more likely to be unemployed. HW children earn 5 percentage points less than WH children, but this gap becomes statistically insignificant when controlling for education.

These results align with existing literature on disparities in educational access \citep{bergman2018education,gaddis2024racial}. While the earnings gap diminishes when controlling for education, this doesn't necessarily indicate an absence of discrimination, as education itself can be influenced by bias. Moreover, educational differences cannot explain the gaps in unemployment, suggesting that discrimination may play a role in labor market outcomes of individuals with Hispanic-sounding last name. Further research is needed to comprehensively understand the earnings gaps between Hispanics and Whites, particularly considering that mothers of HW children have higher education and earnings, which theoretically should lead to better outcomes for their children.
